---
title: "New York State Biodiversity"
subtitle: 
author: "Abigail Kaurin"
institute: "Houghton University"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(xaringanthemer)
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(fig.retina = 3, dpi = 300, fig.width = 6, fig.asp = 0.618, out.width = "80%")
```

```{r echo = FALSE}

knitr::opts_chunk$set(echo = FALSE)

```


```{r load-data, include=FALSE}
# Load your data here

library(tidyverse)
library(dplyr)
library(datasets)
library(readr)

biodiversity <- read.csv("/cloud/project/data/biodiversity-by-county-distribution-of-animals-plants-and-natural-communities-1.csv")

```

```{r}
style_xaringan(
  title_slide_background_image = "img/confetti.jpg")


```


class: center, middle

## What is the variation of biodiversity between NYS counties?
 
---

# Do the ecosystems existing in certain counties affect what species live there?

--

# Do certain counties promote an incline or decline in biodiversity? 

---
class: inverse, center, middle

# Data

---

#Dataset

- I will be using a data set called `Biodiversity-by-County-Distribution-of-Animals-Plants-and-Natural-Communities`

- This data contains `20017` observations, and `12` variables, which are categorized by `county`, `taxonomic group`, `scientific name`, `year last documented`, `NY listing status`, and more.

---

class: inverse, center, middle

#Results

---
#Variation of wildlife in each NYS county

- Which counties are the most diverse?

```{r}
library(dplyr)

biodiversity_count <- biodiversity %>%
  group_by(County) %>%
  summarise(Unique_Species_Count = n_distinct(Scientific.Name))

biodiversity_count <- biodiversity_count %>%
  arrange(desc(Unique_Species_Count))

print(biodiversity_count)

#This shows the distribution of biodiversity by county from highest to lowest.Suffolk, Essex, and Orange are the top three most diverse counties
```



---

# Which counties are the least diverse?

--

```{r}
#This distribution shows biodiversity of counties from lowest bio diverse county to the highest

library(dplyr)

biodiversity_count <- biodiversity %>%
  group_by(County) %>%
  summarise(Unique_Species_Count = n_distinct(Scientific.Name))

biodiversity_count <- biodiversity_count %>%
  arrange(Unique_Species_Count)

print(biodiversity_count)

```


--

- Counties Unknown, "not listed",  Lake Erie Open Waters, Lake Ontario Open Waters,  Atlantic Ocean and Long Island Sound are not applicable.


---

# Accurate table of the least diverse counties:


```{r only counties}

#From here I will remove `Counties Unknown`, `not listed`,	`Lake Erie Open Waters`,	`Lake Ontario Open Waters`, `Atlantic Ocean`, and `Long Island Sound`, as they do not count as official counties.Based on this information, `New York`, `Montgomery`, and `Kings` counties rank with the lowest recorded biodiversity, with Houghton University's Allegany county not far behind them at the 7th lowest position.

biodiversity_count <- biodiversity %>%
  group_by(County) %>%
  summarise(Unique_Species_Count = n_distinct(Scientific.Name))

biodiversity_count <- biodiversity_count %>%
  arrange(Unique_Species_Count)

non_county_entries <- c("Counties Unknown", "not listed", "Lake Erie Open Waters", 
                        "Lake Ontario Open Waters", "Atlantic Ocean and Long Island Sound")

biodiversity_count <- biodiversity_count %>%
  filter(!(County %in% non_county_entries))

print(biodiversity_count)

```


- New York, Montgomery, and Kings counties show the lowest biodiversity.  

- Allegany County is in the 7th lowest position.

---
- Visualization of the biodiversity between counties:

```{r}

library(ggplot2)

biodiversity_count <- biodiversity %>%
  group_by(County) %>%
  summarise(Unique_Species_Count = n_distinct(Scientific.Name))

biodiversity_count <- biodiversity_count %>%
  arrange(Unique_Species_Count) %>%
    filter(!(County %in% non_county_entries))


ggplot(biodiversity_count, aes(x = reorder(County, Unique_Species_Count), y = Unique_Species_Count)) +
  geom_bar(stat = "identity", fill = "grey", width = 0.7) +
  coord_flip() + 
  labs(title = "Distribution of Biodiversity Across NYS Counties",
       x = "New York Counties",
       y = "Unique Species Count") +
  theme_minimal() +
   theme(axis.text.y = element_text(size = 6, hjust = 0))

#Visualization of biodiversity distribution. Based on this information, Suffolk county contains the greatest variability of species by far compared to other counties.***

```

- Suffolk has more than double the amount of unique species that the other counties.

---


# Counties ranked by amount of endangered species:

```{r}

endangered_species <- biodiversity %>%
  filter(NY.Listing.Status == "Endangered" | 
           Federal.Listing.Status == "Endangered" |
           State.Conservation.Rank == "S1")

endangered_count <- endangered_species %>%
  group_by(County) %>%
  summarise(Endangered_Species_Count = n_distinct(Scientific.Name))

endangered_count <- endangered_count %>%
  arrange(desc(Endangered_Species_Count))

print(endangered_count)
#Based on this table, Suffolk, Richmond, and Nassau counties carry the most endangered species in NY. Interestingly Allegany county ranks 60th of counties with endangered species, carrying only 10.

```

- Suffolk, Richmond, and Nassau counties carry the most endangered species in NY. 

- Allegany county ranks 60th of counties with endangered species, carrying only 10.

---

- Visualization of counties carrying endangered species:

```{r endangered-by-county}


endangered_count <- endangered_count %>%
  arrange(desc(Endangered_Species_Count))


ggplot(endangered_count, aes(x = reorder(County, Endangered_Species_Count), 
                             y = Endangered_Species_Count)) +
  geom_bar(stat = "identity", fill = "pink", width = 0.7) +
  coord_flip() +  
  labs(title = "Endangered Species by County",
       x = "New York Counties",
       y = "Number of Endangered Species") +
  theme_minimal() +
   theme(axis.text.y = element_text(size = 5, hjust = 0))

#visualization of endangered species by county.
```

- Suffolk county suprisingly contains the largest amount of endangered species.

---

class: inverse, center middle

# Conclusion

---

#Final Observations:

--

- Difficult to answer the question of what factors specifically lead to there being higher or lower biodiversity, and more or fewer endangered species in these counties.

--

- Only variables that accounted for what animals lived in these counties, not the types of biomes or ecosystems.

--

- Number of variables present made it difficult to display an appealing visualization due to there being so many counties.

---

# Works Cited

Seren Smith, 10 Sept, 2019. "Biodiversity by County - Distribution of Animals, Plants and Natural Communities", https://www.kaggle.com/datasets/jennlsmith/biodiversity-by-county/data

date accessed: 27 April, 2024


New York State, 25 Feb, 2022. "Biodiversity by County - Distribution of Animals, Plants and Natural Communities", State of New York, https://data.ny.gov/Energy-Environment/Biodiversity-by-County-Distribution-of-Animals-Pla/tk82-7km5/about_data

date accessed: 27 April, 2024

